---
title: "blizzardassign"
output: html_document
date: "2024-12-20"
---

Table of Contents:
1. Establish connection to database in PostGreSQL via RStudio
2. Upload raw datasets into RStudio
3. Store datasets into PostGreSQL tables
4. Create intermediate datasets: dim_customer, dim_product, fact_sales
5. Create final dataset: sales_performance

```{r}
install.packages("DBI")
install.packages("RPostgres")

```

1. Establish connection to database via RStudio

```{r}
library(DBI)
library(RPostgres)

# Connection parameters
host <- "localhost"          # server address
port <- 5432                 # PostgreSQL port
dbname <- "blizzardtest"    # database name
user <- "postgres"      # PostgreSQL username
password <- "twixrocks"  # PostgreSQL password

# Establish connection
con <- dbConnect(
  RPostgres::Postgres(),
  host = host,
  port = port,
  dbname = dbname,
  user = user,
  password = password
)
```


2. Upload raw datasets into RStudio

```{r}
# upload csv dataset
orders <- read.csv("/Users/Shared/orders.csv", fileEncoding="UTF-16LE")
products <- read.csv("/Users/Shared/products.csv",fileEncoding="UTF-16LE")
customers <- read.csv("/Users/Shared/customers.csv",fileEncoding="UTF-16LE")

# convert back to numeric and date columns - were uploaded into R as character
orders$total_amount <- as.numeric(orders$total_amount)
orders$order_date <- as.Date(orders$order_date) # converts invalid dates to NA (ex: february 30th)
```

3. Store datasets into PostGreSQL tables

```{r}
# upload data into database
datasets <- list(customers = customers, products = products, orders = orders)
dataset_names <- c("customers", "products", "orders")

# get maximum memory required for allocation of CUSTOMERS variables 
max_customer_id <- max(nchar(customers$customer_id), na.rm = TRUE)
max_customer_name <- max(nchar(customers$customer_name), na.rm = TRUE)
max_region <- max(nchar(customers$region), na.rm = TRUE)
max_segment <- max(nchar(customers$segment), na.rm = TRUE)

# query to create initial CUSTOMERS table
create_table_query <- paste0("CREATE TABLE customers(
                      customer_id VARCHAR(", max_customer_id, ") PRIMARY KEY,
                      customer_name VARCHAR(", max_customer_name, "),
                      region VARCHAR(", max_region, "),
                      segment VARCHAR(", max_segment, "),
                      irrelevant_col_1 VARCHAR(50),
                      irrelevant_col_2 VARCHAR(50),
                      irrelevant_col_3 VARCHAR(50))")
dbExecute(con, create_table_query)

# get maximum memory required for allocation of PRODUCTS variables
max_product_id <- max(nchar(products$product_id), na.rm = TRUE)
max_product_name <- max(nchar(products$product_name), na.rm = TRUE)
max_category <- max(nchar(products$category), na.rm = TRUE)

# query to create initial PRODUCTS table
create_table_query <- paste0("CREATE TABLE products(
                      product_id VARCHAR(", max_product_id, ") PRIMARY KEY,
                      product_name VARCHAR(", max_product_name, "),
                      category VARCHAR(", max_category, "),
                      irrelevant_col_1 VARCHAR(50),
                      irrelevant_col_2 VARCHAR(50),
                      irrelevant_col_3 VARCHAR(50))")
dbExecute(con, create_table_query)

# get maximum memory required for allocation of ORDERS variables
max_customer_id2 <- max(nchar(orders$customer_id), na.rm = TRUE)
max_product_id2 <- max(nchar(orders$product_id), na.rm = TRUE)
max_order_id <- max(nchar(orders$order_id), na.rm = TRUE)
max_total_amount <- max(nchar(as.character(orders$total_amount)), na.rm = TRUE) 

# query to create initial ORDERS table
create_table_query <- paste0("CREATE TABLE orders(
                      order_id VARCHAR(", max_order_id, ") PRIMARY KEY,
                      customer_id VARCHAR(", max_customer_id2, "),
                      order_date DATE,
                      total_amount NUMERIC(", max_total_amount, ", 2),
                      product_id VARCHAR(", max_product_id2, "),
                      irrelevant_col_1 VARCHAR(50),
                      irrelevant_col_2 VARCHAR(50),
                      irrelevant_col_3 VARCHAR(50))")
dbExecute(con, create_table_query)


# upload datasets into SQL tables
for (d in 1:length(datasets)){
  
  # note: csv import doesn't enforce constraints
  dbWriteTable(
    con,
    name = dataset_names[[d]],  
    value = datasets[[d]],
    row.names = FALSE,       
    overwrite = TRUE
  )
}
```

4. Create intermediate datasets

a) dim_customer
```{r}
# create dim_customer table
### setting customer_id as primary key will result in error if trying to upload duplicate customer ID
### set memory as maximum length found in column within dataset
create_table_query <- paste0("CREATE TABLE dim_customer(
                      customer_id VARCHAR(", max_customer_id, ") PRIMARY KEY,
                      customer_name VARCHAR(", max_customer_name, "),
                      region VARCHAR(", max_region, "),
                      segment VARCHAR(", max_segment, "))")
dbExecute(con, create_table_query)

# insert data into dim_customer
### de-duplicate customer IDs by adding primary key constraint
### remove all white space from customer_id, and remove trailing/leading spaces for remaining varchar variables
### convert all varchar variables to uppercase for consistency
### remove empty/null customer_ids or those equal to NA
### remove duplicate rows, although they shouldn't appear
### set region to NULL if not in N/E/S/W
create_table_query <- "INSERT INTO dim_customer
                      SELECT DISTINCT NULLIF(UPPER(REPLACE(customer_id, ' ', '')), '') AS customer_id, 
                      TRIM(UPPER(customer_name)) AS customer_name, 
                      CASE 
                          WHEN TRIM(UPPER(region))  NOT IN ('NORTH', 'EAST', 'SOUTH', 'WEST') THEN NULL
                          ELSE TRIM(UPPER(region)) 
                      END AS region,
                      TRIM(UPPER(segment)) AS segment
                      FROM customers
                      WHERE NULLIF(UPPER(REPLACE(customer_id, ' ', '')), '') != 'N/A' AND NULLIF(UPPER(REPLACE(customer_id, ' ', '')), '') IS NOT NULL"
dbExecute(con, create_table_query)
```

b) dim_product
```{r}
# create dim_product table
### setting product_id as primary key will result in error if trying to upload duplicate product ID
### set memory as maximum length found in column within dataset
create_table_query <- paste0("CREATE TABLE dim_product(
                      product_id VARCHAR(", max_product_id, ") PRIMARY KEY,
                      product_name VARCHAR(", max_product_name, "),
                      category VARCHAR(", max_category, "))")
dbExecute(con, create_table_query)

# insert data into dim_product
### de-duplicate product IDs by adding primary key constraint
### remove all white space from product_id, and remove trailing/leading spaces for remaining varchar variables
### convert all varchar variables to uppercase for consistency
### remove empty/null product_ids or those equal to NA
### remove duplicate rows, although they shouldn't appear
create_table_query <- "INSERT INTO dim_product
                      SELECT DISTINCT NULLIF(UPPER(REPLACE(product_id, ' ', '')), '') AS product_id, 
                      TRIM(UPPER(product_name)) AS product_name, 
                      TRIM(UPPER(category)) AS category
                      FROM products
                      WHERE NULLIF(UPPER(REPLACE(product_id, ' ', '')), '') != 'N/A' AND NULLIF(UPPER(REPLACE(product_id, ' ', '')), '') IS NOT NULL"
dbExecute(con, create_table_query)
```

c) fact_sales
```{r}
# create fact_sales
### setting product_id as primary key will result in error if trying to upload duplicate product ID
### set memory as maximum length found in column within dataset; want 2 decimal places for total_amount
create_table_query <- paste0("CREATE TABLE fact_sales(
                      order_id VARCHAR(", max_order_id, ") PRIMARY KEY,
                      customer_id VARCHAR(", max_customer_id2, "),
                      order_date DATE,
                      total_amount NUMERIC(", max_total_amount, ", 2),
                      product_id VARCHAR(", max_product_id2, "),
                      customer_name VARCHAR(", max_customer_name, "),
                      region VARCHAR(", max_region, "),
                      segment VARCHAR(", max_segment, "),
                      product_name VARCHAR(", max_product_name, "),
                      category VARCHAR(", max_category, "))")
dbExecute(con, create_table_query)

# insert data into fact_sales
### replace varchar variables with NULL if they are empty after removing all whitespace (most done in prev tables)
### if total_amount is NULL then set to 0
### remove duplicate rows, although they shouldn't appear
### invalid dates are set to NULL/NA already, so no adjustment here
create_table_query <- "INSERT INTO fact_sales
                      SELECT DISTINCT NULLIF(UPPER(REPLACE(aa.order_id, ' ', '')), '') AS order_id,
                      NULLIF(UPPER(REPLACE(aa.customer_id, ' ', '')), '') AS customer_id, 
                      aa.order_date, 
                      COALESCE(aa.total_amount, 0) AS total_amount,
                      NULLIF(UPPER(REPLACE(aa.product_id, ' ', '')), '') AS product_id,
                      NULLIF(bb.customer_name, '') AS customer_name,
                      NULLIF(bb.region, '') AS region,
                      NULLIF(bb.segment,'') AS segment,
                      NULLIF(cc.product_name, '') AS product_name,
                      NULLIF(cc.category, '') AS category
                      FROM orders AS aa
                      LEFT JOIN dim_customer AS bb
                      ON NULLIF(UPPER(REPLACE(aa.customer_id, ' ', '')), '') = bb.customer_id
                      LEFT JOIN dim_product AS cc
                      ON NULLIF(UPPER(REPLACE(aa.product_id, ' ', '')), '') = cc.product_id
                      WHERE NULLIF(UPPER(REPLACE(aa.order_id, ' ', '')), '') != 'N/A' AND NULLIF(UPPER(REPLACE(aa.order_id, ' ', '')), '') IS NOT NULL"
dbExecute(con, create_table_query)
```

5. Create final dataset, sales_performance

```{r}
# each row is the amount of money an individual customer (determined by customer_id) spent within a month on a certain product
create_table_query <- "CREATE TABLE sales_performance AS
                       SELECT customer_name, 
                       region, 
                       segment, 
                       product_name, 
                       category, 
                       EXTRACT(MONTH FROM order_date) AS month, 
                       SUM(total_amount) AS total_revenue
                       FROM fact_sales
                       GROUP BY month, customer_id, customer_name, region, segment, product_name, category
                       ORDER BY month DESC NULLS LAST, total_revenue DESC "
dbExecute(con, create_table_query)
```

```{r}
query <- "SELECT * FROM sales_performance"
result <- dbGetQuery(con, query)
result
```




